Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-23 paper/paper.tex
fix misc mistakes

In Danvy's paper \textit{Defunctionalized Interpreters for Programming
  Languages}, he mentions that for deterministic languages, a reduction semantics
is a structural operational semantics in continuation style, where the reduction
context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
This poses a problem, as the underlying semantics of AAM is fundamentally nondeterministic.
But the evaluation context (i.e., the @Frame@ in our program) is \emph{not}
nondeterministic, and the worklist @todo@ actually implicitly handles the nondeterminism.
Thus, if we simply refunctionalize the frames to higher-order functions, 
it does not help us to move toward abstract definitional interpreters.

@@

The above code shows the skeleton of @drive_step@.
At the end of the first case of pattern matching, we invoke @drive_step@ recursively 
\footnote{The code for constructing new environments, stores, and frames is elided for 
simplicity of presentation.}.
The code for the second case is replaced entirely by a function call to @continue@.
The shape of an interpreter in continuation-passing style has emerged.
%The astute reader will notice a vague shape of interpreter in continuation style has emerged.

@@

Readers familiar with functional programming may think of @fold@ at this point; indeed,
it is a variant implementation of @fold@ in continuation style. But here we have two
continuations, one for the next element in the collection, and one for the eventual result.
This is sometimes referred to as an extended continuation-passing style (ECPS)
that function @k@ is a normal continuation and @m@ is a meta-continuation 
\cite{Danvy:1990:AC:91556.91622}. \\

@@

The function @analyze@ is also changed by adding a @reset@ operator around the function 
call to @aeval@ to set the delimiter. The afterwards updating on cache becomes sequential.

@@

Now we have arrived the end of this series of transformations. Starting from an 
small-step abstract abstract machine, eventually we obtain an big-step definitional
interpreter with pushdown control-flows and written in direct-style.

@@

For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows
a reduction semantics is a structural operational semantics in continuation style,
where the reduction context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
% Danvy also shows in nondeterministic languages ...
Our work further shows that for nondeterministic languages, there is another layer of continuations
that controls the nondeterministic choices. After identifying and explicitly exposing 
(Section~\ref{linear}) that layer of nondeterministic continuations, the whole program can 
then be refunctionalized to extended continuation-passing style.

@@

\citeauthor{gilray2016pushdown} further proposed a polyvariant continuation-addresses allocator
for small-step AAM to achieve pushdown analysis \cite{gilray2016pushdown}.
This method is both simple to implement and computationally inexpensive and so called Pushdown 
for Free (P4F). Based on the AAM we presented in Section \ref{aam}, the only changes in the 
code required is not only keeping track of the entry-point expression of callee, but also 
holding the target environment when allocating a continuation addresses.
No other pieces of code would need to be modified.
Notably, this change only causes constant-factor increase in time complexity of
the analysis if the store is widened.

@@

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work.
The difference on the surface is that ours works for A-Normal Form $\lambda$-calculus,
instead of plain $\lambda$-calculus, although our work can also be easily adopted to handle a plain
$\lambda$-calculus.
But instead of directly showing the final form of abstract definitional interpreters,
we reveal that refunctionalization plays an important role for inheriting the
stack from the metalanguage.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written direct-style with
delimited control operators. 
We use the delimited control operators @shift@ and @reset@ that are implemented in Scala 
\cite{rompf2009implementing}. The correspondence of delimited control (as well as continuations) 
and monads are well-known \cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.

>>>>

In Danvy's paper \textit{Defunctionalized Interpreters for Programming
  Languages}, he mentions that for deterministic languages, a reduction semantics
is a structural operational semantics in continuation-passing style, where the reduction
context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
This poses a problem, as the underlying semantics of AAM is fundamentally nondeterministic.
But the evaluation context (i.e., the @Frame@ in our program) is \emph{not}
nondeterministic, and the worklist @todo@ actually implicitly handles the nondeterminism.
Thus, if we simply refunctionalize the frames to higher-order functions, 
it does not help us to move toward abstract definitional interpreters.

@@

The above code shows the skeleton of @drive_step@.
At the end of the first case of pattern matching, we invoke @drive_step@ recursively 
\footnote{The code for constructing new environments, stores, and frames is elided for 
simplicity of presentation.}.
The code for the second case is replaced entirely by a function call to @continue@.
The shape of an interpreter in continuation-passing style has emerged.

@@

Readers familiar with functional programming may think of @fold@ at this point; indeed,
it is a variant implementation of @fold@ in continuation-passing style. But here we have two
continuations, one for the next element in the collection, and one for the eventual result.
This is sometimes referred to as an extended continuation-passing style (ECPS)
that function @k@ is a normal continuation and @m@ is a meta-continuation 
\cite{Danvy:1990:AC:91556.91622}. \\

@@

The function @analyze@ is also changed by calling a @reset@ operator around the function 
call of @aeval@ to set the delimiter. The afterwards updating on cache becomes sequential.

@@

Now we have arrived the end of this series of transformations. Starting from a
small-step abstract abstract machine, eventually we obtain a big-step definitional
interpreter with pushdown control-flows and written in direct-style.

@@

For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows
a reduction semantics is a structural operational semantics in continuation-passing style,
where the reduction context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
% Danvy also shows in nondeterministic languages ...
Our work further shows that for nondeterministic languages, there is another layer of continuations
that controls the nondeterministic choices. After identifying and explicitly exposing 
(Section~\ref{linear}) that layer of nondeterministic continuations, the whole program can 
then be refunctionalized to extended continuation-passing style.

@@

\citeauthor{gilray2016pushdown} further proposed a polyvariant continuation-addresses allocator
for small-step AAM to achieve pushdown analysis \cite{gilray2016pushdown}.
This method is both simple to implement and computationally inexpensive and so called Pushdown 
for Free (P4F). Based on the AAM we presented in Section \ref{aam}, the only changes in the 
code required is not only keeping track of the entry-point expression of the callee, but also 
holding the target environment when allocating a continuation addresses.
No other pieces of code would need to be modified.
Notably, this change only causes constant-factor increase in time complexity of
the analysis if the store is widened.

@@

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work.
The difference on the surface is that ours works for A-Normal Form $\lambda$-calculus,
instead of plain $\lambda$-calculus, although our work can also be easily adapted to handle a plain
$\lambda$-calculus.
But instead of directly showing the final form of abstract definitional interpreters,
we reveal that refunctionalization plays an important role for inheriting the
stack from the metalanguage.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written direct-style with
delimited control operators. 
We use the delimited control operators @shift@ and @reset@ that are implemented in Scala 
\cite{rompf2009implementing}. The correspondence of delimited control (as well as continuations) 
and monads are well-known \cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.
